Outline flowCore paper

### 2008-04-30 paper/outline.tex
illustration and citation
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31442 bc3139a8-67e5-0310-9ffc-ced21a209358

\documentclass[12pt]{article}
\usepackage{url}
\usepackage[authoryear,round]{natbib}

@@

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information.  Moreover, it is one of the very first software
implementations of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing work flows and to communicate with
virtually every other FCM tool that adheres to the proposed
standard. In addition, the software is platform independent, running
on various operating systems including Windows, Mac OS X and
Linux/Unix.

@@

\subsection*{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way:
namely, more or less ad hoc spreadsheets that lack a controlled
structure. In general, there is no native ordering to a flowSet, which
in a way reflects the still predominant application of tube-based
experiments rather than experiments following a plate-design. However,
the flowSet structure can be readily extended to also accommodate such
spatial information \citep{(plateCore)}

\textit{Figure: A diagram of the data structures and the basic
  operations (Figure1.ppt)}

\section*{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and access. Third
party methods implemented by others can act on their own as
first-class citizens in the analysis framework, without breaking the
workflow or the basic infrastructure. This design allows for the
straight forward extension of flowCore's capabilities and already
fostered the development of a number of valuable
improvements. \citep{lo2008agf} (+plateCore).

\subsection*{Transformation and compensation}
Proper transformation have been proven beneficial for both data
visualization and modelling \textit{cite flowClust paper}. All the
major transformations that are routinely used in flow cytometry
analysis have been implemented in flowCore. Furthermore, the design of
the R language makes it easy to define arbitrary functions to apply to
the data of individual flowFrames or entrire flowSets,
respectively. Compensation is available for both flowFrames and
flowSets, and the software also offers functionality to compute
spillover or compensation matrices from a set of appropriate
compensation samples.

@@

\section*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data
analysis. The flowViz package provides sophisticated data
visualization tools, that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
web-based reports of QA results. Most of flowUtil's content deals with
data import and output from and to flowCore and the flow-cytometry
specific standard markup languages. Finally, flowStats provides
statistical methods that are relevant in the context of flow cytometry
data analysis.

\textit{A diagram of the packages and what they offer}

@@

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flow frame or a complete flow set
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarily, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

Static gating for all samples in a high-throughput flow cytometry
experiment is often not an option, since the measured variables tend
to vary between different treatments or over time. Automated or
data-driven gating tries to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\cite{lo2008agf}. One of the automated gating methods implemented in
flowCore is based on identifying areas of significant curvature in a
kernel density estimate of the data (\ref{Wandetal}). Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in either the one ore the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it(\ref{Wandetal}). The
modular design of R, Bioconductor and flowCore allows to easily
integrate these existing solutions into our framework. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all funtionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{url}
\usepackage[authoryear,round]{natbib}

@@

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}).  Moreover, it is one of the very first software
implementations of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing work flows and to communicate with
virtually every other FCM tool that adheres to the proposed
standard. In addition, the software is platform independent, running
on various operating systems including Windows, Mac OS X and
Linux/Unix.

@@

\subsection*{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way:
namely, more or less ad hoc spreadsheets that lack a controlled
structure. In general, there is no native ordering to a flowSet, which
in a way reflects the still predominant application of tube-based
experiments rather than experiments following a plate-design. However,
the flowSet structure can be readily extended to also accommodate such
spatial information [citing plateCore - Any reference here Perry, Errol?]

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{The flowCore frame work.For each experiement,
  FCS files, phenotypic and meta data can be store in a flowSet. Each
  flowFrame of a flowSet corresponds to one FCS file.  Any basic
  operation (compensate, filter, ...) can be applied on a flowSet or a
  flowFrame.Different summary output can be generated with flowCore or
  the associated packages.}}
\end{figure}

\section*{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and access. Third
party methods implemented by others can act on their own as
first-class citizens in the analysis framework, without breaking the
workflow or the basic infrastructure. This design allows for the
straight forward extension of flowCore's capabilities and already
fostered the development of a number of valuable
improvements \citep{lo2008agf, sarkar2008ufv}.

\subsection*{Transformation and compensation}
Proper transformation have been proven beneficial for both data
visualization and modelling \citep{lo2008agf}. All the
major transformations that are routinely used in flow cytometry
analysis have been implemented in flowCore. Furthermore, the design of
the R language makes it easy to define arbitrary functions to apply to
the data of individual flowFrames or entrire flowSets,
respectively. Compensation is available for both flowFrames and
flowSets, and the software also offers functionality to compute
spillover or compensation matrices from a set of appropriate
compensation samples.

@@

\section*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data analysis
(Figure~\ref{fig2:flowWheel}). The flowViz package provides
sophisticated data visualization tools, that make use of multivariate
trellis plotting. These functions can be used to quickly generate
customized plots for extended cytometry data sets for both direct data
inspection and quality control. The flowQ package offers more advanced
quality assurance (QA) methodology and a framework to create
interactive web-based reports of QA results. Most of flowUtil's
content deals with data import and output from and to flowCore and the
flow-cytometry specific standard markup languages. Finally, flowStats
provides statistical methods that are relevant in the context of flow
cytometry data analysis.

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
\caption{\label{fig2:flowWheel}{The flowCore Suite: visualization and analysis
  packages relying on the flowCore data structure.  Dedicated packages
  have been developped using the flowCore data structure for
  manipulating the experiemental data: flowViz is a data visualization
  package; flowQ helps at data quality assessment; flowUtils provides
  xml parser, flowClust proposes population 
  filtration using model-based clustering}}
\end{figure}

@@

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flow frame or a complete flow set
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarily, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package \citep{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

Static gating for all samples in a high-throughput flow cytometry
experiment is often not an option, since the measured variables tend
to vary between different treatments or over time. Automated or
data-driven gating tries to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\citep{lo2008agf}. One of the automated gating methods implemented in
flowCore is based on identifying areas of significant curvature in a
kernel density estimate of the data
\citep{wand2008,wand2008_2}. Assuming that the regions of interest are
of high density, the software is able to reliably detect them in
either the one ore the two-dimensional density landscape. Kernel
density estimation is a well-known problem in statistical computing,
and a lot of effort has been invested in the development of good
software to address it \citep{wand2008,wand2008_2}. The modular design
of R, Bioconductor and flowCore allows to easily integrate these
existing solutions into our framework. Instead of re-writing existing
code, we are able to include it via the well tested distribution
mechanism provided by R's software package system. This process is
bi-directional, and all funtionality implemented in flowCore is
available to other package authors, as we have seen with the
afore-mentioned flowClust package.
